---
title: "Stat528_Group1_Project"
author: "Group1"
date: "November 15, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Overview

For our group assignment, we have chose to look at a data set from an Animal Shelter's in Dallas, TX. Animal shelters help animals find new homes and maintain a safe city. We believe this report can be used to help this shelter save money and operate more efficiently. In addition, this data set can be used in various ways such as marketing, keeping appropriate inventory of pet supplies on hand, and staffing necessary resources at certain times.

The Dallas Animal Shelter data set can be found on the [TidyTuesday](https://github.com/rfordatascience/tidytuesday) project website at the following link: [week18_dallas_animals.xlsx](https://github.com/rfordatascience/tidytuesday/blob/master/data/2018-07-31/week18_dallas_animals.xlsx)

Each data point represents a different animal that entered the shelter identified by an `andimal_id`. Each observation includes the type of animal, the breed of the animal, the `intake_type` and `outcome_type`, and dates of these events. In addition, there is information about what condition the animal was in at the time of intake and at the time of the outcome.

Our intentions with this data set is to answer a multitude of different questions that could help the animal shelter in the future. We think the following questions can help the animal shelter optimize a plan for incoming pets and try to maximize the opportunity for a pet to get adopted:

* Is there any spikes for adoptions and intake? And can we link them to any reasons?
* How long are animals typically at the shelter? How long for different outcomes?
* Which animals are most likely to be adopted?
* Which animals/breed  are adopted the quickest?
* Does a particular `intake_type` or condition lead to a faster turnaround?

## Data Description

To read in this data, we have downloaded the file from the above link and created a project to easily read in the file. This `R-Markdown` is part of a package. If you un-zip and open the package, you should be able to run this code easily. One thing to note, the following packages were used to create this report and you will need to `install.packages` before running the code:

* `readxl` - to read in excel files.
* `scales` - for adjusting graphics legends and axes.
* `lubridate` - fast and user friendly way to work with dates.
* `rvest` - for scraping information off of the web.
* `RColorBrewer` - for changing graph colors
* `tidyverse` - a collection of packages to manipulate and graph data.


```{r library bank, echo=FALSE}
#====library bank====
library(readxl)
library(scales, warn.conflicts = F)
library(lubridate,warn.conflicts = F)
library(rvest, warn.conflicts = F, quietly = T)
library(RColorBrewer)
suppressPackageStartupMessages(library(tidyverse,
                                       warn.conflicts = F))
```

To get started, we will walk you through our analysis process. One thing we noticed immediately with this data set was that there were two tabs on the Excel sheet.
The first tab was raw, while the second was a sample of the raw. To ensure we had all the data possible we did a quick comparison of the two sets using `left_joining`  the `raw` tab's column names to the `simple` tab's column names and making sure there was no `NA` values. We then ensured the lengths of each tab were the same using `nrow`.

```{r col and row check, echo=FALSE}
#check that no columns in raw tab are in the simple tab.
suppressWarnings(data.frame(colname_simple = as.character(colnames(read_excel("week18_dallas_animals.xlsx", sheet = "simple")))) %>%
  left_join(data.frame(colname_simple = as.character(colnames(read_excel("week18_dallas_animals.xlsx", sheet = "raw"))),
                       colname_raw = as.character(colnames(read_excel("week18_dallas_animals.xlsx", sheet = "raw")))),
            by = "colname_simple"))

#check that rows are the same length
nrow(read_excel("week18_dallas_animals.xlsx", sheet = "simple")) == nrow(read_excel("week18_dallas_animals.xlsx", sheet = "raw"))
```

This proves that the `simple` tab is just a subset of the `raw` tab. Since the `raw` data isn’t overly large, we will just use the `raw` tab and filter out columns that really won't help us. This way we will avoid missing any conclusions that can be drawn from the `raw` tab information that cannot be drawn from the `simple` tab. Belwo is a `glimpse` of the Dallas Animal Shelter we will be using:

```{r read in file, echo=FALSE}
dallas_animals <- read_excel("week18_dallas_animals.xlsx", sheet = "raw", na = "NA")

dallas_animals <- dallas_animals %>%
  select(-c(kennel_status, tag_type, activity_number, activity_sequence, source_id,
            census_tract, receipt_number, impound_number, service_request_number,
            chip_status, year)) %>%
  rename(year = mo_year)

glimpse(dallas_animals)

```

## Seasonality for Intake and Adoption

To get a better understanding of our data we will use a time series analysis for the number of intakes and the number of outcomes. This will give us an understanding of the time frame of our data and see if we can unveil any possible seasonality in the Dallas Animal Shelter data for intake and adoption. In addition, we will look at the types of intakes and outcomes.

### Intake Analysis

There is **`r format(nrow(dallas_animals), big.mark = ",", scientific = FALSE, trim = F)`** animals which the Dallas Animal Shelter has taken in. The time series graph below which gives us an idea of the time frame of our data and any obvious trends. To create this graph, we will use `geom_line` to create the line graph and then add `geom_smooth` to capture any trends.

```{r time series intake, echo=TRUE}
dallas_animals %>% count(intake_date) %>%
  ggplot(aes(x = intake_date, y = n)) +
  geom_line() +
  geom_smooth(method = 'auto') +
  ggtitle("Intake Time Series") +
  xlab("Intake Date") +
  ylab("Number of Animals") +
  theme(plot.title = element_text(face="bold")) +
  theme_bw()
```

This graph shows us that we have one years worth data from the Dallas Animal Shelter. It also shows us that we could possibly have some cyclical data. The smoothing line dips in the winter months and then spikes in the spring and summer months. However, one years worth of data is not really sufficient for identifying seasonality. Another thing to point out is that there are large spikes and dips in our data. If this is the case when we investigate adoption, it will be worth while to take a look at days around the holiday's to see if there are any spikes around the holiday's, especially Christmas.

Right now, we will look at why animals are being administered to the Dallas Animal Shelter. The graph below shows different `intake_types` ordered from most common `intake_type` to least common.

```{r intake types, echo=TRUE}
dallas_animals %>% count(intake_type) %>%   #count the number of animals
  ggplot(aes(x = reorder(intake_type, n), y = n)) +
  geom_bar(stat = "identity") +             #create a bar graph
  geom_text(aes(label = comma(n)),          #add labels to the graph with a 1,000's comma
            position = position_dodge(0.9),
            vjust = 0.1,
            hjust = -0.2) +
  ggtitle("Intake Types") +
  xlab("Intake Type") +
  ylab("Number of Animals") +
  scale_y_continuous(labels = comma, limits = c(0, 25000)) +
  theme(plot.title = element_text(face="bold")) +
  theme_bw() +
  coord_flip()
```

We can see that `Owner Surrender` and `Stray` is high on the list. To look closer at these predominate `intake_type`s, we will add a column to the data set called `intake_day` which gives the day of the week of the `intake_date`. Then we can `filter` down to `intake_type`s equal to `STRAY` or `OWNER SURRENDER`. Then we can make a graph to see if there are days of the week where an owner is more likely to surrender their pet or it is more likely someone will call in a stray.

```{r day of week, echo=FALSE}
dallas_animals <- dallas_animals %>% mutate(intake_day = wday(intake_date, label = TRUE))
```

```{r top intake by day, echo=TRUE}
dallas_animals %>% filter(intake_type == "STRAY" | intake_type == "OWNER SURRENDER") %>%
  ggplot(aes(x = intake_day, fill = intake_type)) +
  geom_bar() +
  facet_wrap( ~ intake_type) +
  ggtitle("Intake Type by Day") +
  xlab("Intake Day") +
  ylab("Number of Animals") +
  labs(fill = "Intake Type") +
  scale_y_continuous(labels = comma) +
  theme_bw() +
  theme(plot.title = element_text(face="bold")) +
  scale_fill_brewer(palette="Set1")
```

There are a couple things we can identify from this graph. One is that intake's are slow on Sundays. After reviewing the Dallas Animal Shelter hours online, the hours are shorter on Sunday, but they seem to have reasonalbe hours every day of the week (assuming I am looking at the right shelter). Another interesting conclusion from this graph, is that there is a larger number or strays found on Wednesdays.

Lastly, for the intake section we will look at what types of animals the Dallas Animal Shelter is taking in.

```{r animal types intake, echo=TRUE}
dallas_animals %>% count(animal_type) %>%   #count the number of animals
  ggplot(aes(x = animal_type, y = n)) +
  geom_bar(stat = "identity") +             #create a bar graph
  geom_text(aes(label = comma(n)),          #add labels to the graph with a 1,000's comma
            position = position_dodge(0.9),
            vjust = -0.3) +
  ggtitle("Animal Intake Types") +
  xlab("Animal Type") +
  ylab("Number of Animals") +
  scale_y_continuous(labels = comma) +
  theme(plot.title = element_text(face="bold")) +
  theme_bw()
```

As expected, there are way more dogs than anything else. Because of the large percentage of dogs compared to every other animal type, it will be worth taking a closer look at dogs specifically. Before we get indepth with our dog analysis, we should finish our analysis on the collection of animals. That way, if we add any variables we will capture those variables in the dog analysis. For now, we can see what types of dog breeds the Dallas Animal Shelter is taking in and why.

Because there are so many dog breed types, we will lump the 20% least common types of dog breeds brought into the Dallas Animal Shelter as `OTHER`. In other words, we will look at the top 80% most common types of dog breeds by name and then lump all other dog breed types into one category.
To do this, we will create a new data frame called `dog_breed` and then add a cummulative percent with the function `cumsum`. With an `ifelse` statement we denote dog breed above 80% in the cummulative column as `OTHER` in a new column `new_breed` and then `left_join` that back into our data frame.

```{r dog breeds, echo=TRUE}
dog_breed <- dallas_animals %>%
  filter(animal_type == "DOG") %>%
  count(animal_breed) %>%
  rename(count_dogs = n) %>%
  arrange(desc(count_dogs)) %>%
  mutate(percent_dogs = round(count_dogs/sum(count_dogs), 2),
         cum_percent = cumsum(percent_dogs),
         new_breed = ifelse(cum_percent > 0.8, "OTHER", animal_breed))

dallas_animals <- dallas_animals %>%
  left_join(dog_breed %>%
              select(animal_breed, new_breed),
            by = "animal_breed")
```



```{r dog breed types, echo=TRUE}
dallas_animals %>%
  filter(animal_type == "DOG") %>%
  count(new_breed) %>%   #count the number of animals
  ggplot(aes(x = reorder(new_breed, n), y = n)) +
  geom_bar(stat = "identity") +             #create a bar graph
  geom_text(aes(label = comma(n)),          #add labels to the graph with a 1,000's comma
            position = position_dodge(0.9),
            vjust = 0.1,
            hjust = -0.2) +
  ggtitle("Count of Dog Breeds") +
  xlab("Dog Breed") +
  ylab("Number of Animals") +
  scale_y_continuous(labels = comma, limits = c(0, 6500)) +
  theme(plot.title = element_text(face="bold")) +
  theme_bw() +
  coord_flip()
```




From there we can create a graphic which tells us the how the different types of dog breeds come into the shelter.


```{r dog breed intake, echo=TRUE}
dallas_animals %>%
  filter(animal_type == "DOG") %>%
  count(intake_type, new_breed) %>%
  ggplot(aes(x = reorder(intake_type, n),
             y = reorder(new_breed, n),
             fill = n)) +
  geom_tile() +
  ggtitle("Dog Breed Intake Types") +
  xlab("Intake Type") +
  ylab("Animal Breed") +
  labs(fill='Count of Dogs') +
  theme_bw() +
  theme(plot.title = element_text(face="bold"),
        axis.text.x = element_text(angle = 45)) +
  scale_fill_gradient(low = "white", high = "steelblue")
```
### Outcome Analysis

Next we will do the same type of analysis on outcomes. First, we will do a time series analysis on the `outcome_date`.

```{r time series outcome, echo=TRUE}
dallas_animals %>% filter(!is.na(outcome_date)) %>% count(outcome_date) %>%
  ggplot(aes(x = outcome_date, y = n)) +
  geom_line() +
  geom_smooth(method = 'auto') +
  ggtitle("Outcome Time Series") +
  xlab("Outcome Date") +
  ylab("Number of Animals") +
  theme(plot.title = element_text(face="bold")) +
  theme_bw()
```

This graph shows us the same as our intake data, the smoothing line dips in the winter months and then spikes in the spring and summer months. Next, we will look at the types of outcomes there are.

```{r outcome types, echo=TRUE}
dallas_animals %>% count(outcome_type) %>%   #count the number of animals
  ggplot(aes(x = reorder(outcome_type, n), y = n)) +
  geom_bar(stat = "identity") +             #create a bar graph
  geom_text(aes(label = comma(n)),          #add labels to the graph with a 1,000's comma
            position = position_dodge(0.9),
            vjust = 0.1,
            hjust = -0.2) +
  ggtitle("Outcome Types") +
  xlab("Outcome Type") +
  ylab("Number of Animals") +
  scale_y_continuous(labels = comma, limits = c(0, 12500)) +
  theme(plot.title = element_text(face="bold")) +
  theme_bw() +
  coord_flip()
```


```{r top intake by day, echo=TRUE}
dallas_animals %>% filter(outcome_type == "ADOPTION" | outcome_type == "EUTHANIZED" | outcome_type == "TRANSFER" | outcome_type == "RETURNED TO OWNER") %>%
  ggplot(aes(x = intake_day, fill = outcome_type)) +
  geom_bar() +
  facet_wrap( ~ intake_type) +
  ggtitle("Intake Type by Day") +
  xlab("Intake Day") +
  ylab("Number of Animals") +
  labs(fill = "Intake Type") +
  scale_y_continuous(labels = comma) +
  theme_bw() +
  theme(plot.title = element_text(face="bold")) +
  scale_fill_brewer(palette="Set1")
```

## Time at Dallas Animal Shelter

An animal shelter obviously houses animals, but sadly, only has room for so many animals. To understand how long animals are in the shelter, we will add a variable called `total_time` which indicates how many days the animal was in the shelter by using `datediff` on `intake_date` and `outcome_date`.

```{r time in shelter, echo=TRUE}
dallas_animals <- dallas_animals %>%
  mutate(total_time = difftime(outcome_date, intake_date, units = "days"))
```

The first thing we can do is look at the average amount of time an animal spends at the Dallas Animal Shelter by taking the mean of the variable `total_time` which we just added. Average Time at Dallas Animal Shelter = `r round(mean(dallas_animals$total_time, na.rm = T), 2)`

Next, we can look at the average time for different `outcome_types`. We will need to use `group_by` on `outcome_type` and `summarize` to get the mean.

```{r mean outcome type, echo=TRUE}
dallas_animals %>%
  group_by(outcome_type) %>%
  summarize(avg_time = round(mean(total_time, na.rm = T), 2)) %>%
  arrange(desc(avg_time))
```

This tells us that if there is an issue with a report, the animal will likely be there a longer time.

One thing we can do to take a closer look at how long an animal spends at the Dallas Animal Shelter is to include `intake_condition` to our grouping.

```{r intake condition and time spent, echo=TRUE}
# dallas_animals %>%
#   group_by(intake_condition, outcome_type) %>%
#   summarize(avg_time = round(as.numeric(mean(total_time, na.rm = T)), 2)) %>%
#   ggplot(aes(x = reorder(outcome_type, avg_time),
#              y = reorder(intake_condition,
#                          c("HEALTHY",
#                            "TREATABLE REHABILITABLE NON-CONTAGIOUS",
#                            "TREATABLE REHABILITABLE CONTAGIOUS",
#                            "TREATABLE MANAGEABLE NON-CONTAGIOUS",
#                            "TREATABLE REHABILITABLE CONTAGIOUS",
#                            "UNHEALTHY UNTREATABLE NON-CONTAGIOUS",
#                            "UNHEALTHY UNTREATABLE CONTAGIOUS")),
#              fill = avg_time)) +
#   geom_tile() + scale_fill_gradient(low = "white", high = "steelblue") +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Plans for Further work:

We will dive deeper into time spent at the Dallas Animal Shelter answering questions like "what type of animal is adopted faster?", "what time of intake is more likely to get adopted?"
Because of the majority of the data is dog related, we will dive further into the dogs and dog breeds. Looking at "does intake type correlate to time spent or adoption rate?"and "which breeds are adopted the quickest?" 
 
## Dog Data Analysis
To take a closer look at the dog data, we will save a data set called `dallas_dogs` which contains only the animal type `DOG`.

```{r dog data, echo=TRUE}
dallas_dogs <- dallas_animals %>% filter(animal_type == "DOG")
```

With the dog data, we can take a look at what kinds of breeds the Dallas Animal Shelter is taking in. Since there is so many dog breeds, we will filter down to the top 10 types of dog breeds that Dallas Animal Shelter is taking in.

```{r dog breeds, echo=TRUE}
dallas_dogs %>%
  count(animal_breed) %>%
  top_n(n = 10, wt = n) %>%
  ggplot(aes(x = reorder(animal_breed, -n), y = n)) +
  geom_bar(stat = "identity") +             #create a bar graph
  geom_text(aes(label = comma(n)),          #add labels to the graph with a 1,000's comma
            position = position_dodge(0.9),
            vjust = -0.1) +
  ggtitle("Top 10 Dog Breed Intake") +
  xlab("Animal Breed") +
  ylab("Number of Animals") +
  theme(plot.title = element_text(face="bold"),
        axis.text.x = element_text(angle = 90))
```



```{r dog breed outcomes, echo=TRUE}
dallas_animals %>%
  filter(animal_type == "DOG") %>%
  count(outcome_type, new_breed) %>%
  ggplot(aes(x = reorder(outcome_type, n),
             y = reorder(new_breed, n),
             fill = n)) +
  geom_tile() +
  ggtitle("Dog Breed Outcomes Types") +
  xlab("Intake Type") +
  ylab("Animal Breed") +
  labs(fill='Count of Dogs') +
  theme_bw() +
  theme(plot.title = element_text(face="bold"),
        axis.text.x = element_text(angle = 45)) +
  scale_fill_gradient(low = "white", high = "steelblue")
```
