---
title: "Dallas Animal Shelter"
author: "Alex Brown"
date: "November 15, 2018"
output:
  github_document:
    toc: true
    number_sections: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project Overview

For this project, I have chose to look at a data set from an Animal Shelter's in Dallas, TX. Animal shelters help animals find new homes and maintain a safe city. I believe this report can be used to help this shelter save money and operate more efficiently. In addition, this data set can be used in various ways such as marketing, keeping appropriate inventory of pet supplies on hand, and staffing necessary resources at certain times.

The Dallas Animal Shelter data set can be found on the [TidyTuesday](https://github.com/rfordatascience/tidytuesday) project website at the following link: [week18_dallas_animals.xlsx](https://github.com/rfordatascience/tidytuesday/blob/master/data/2018-07-31/week18_dallas_animals.xlsx)

Each data point represents a different animal that entered the shelter identified by an `andimal_id`. Each observation includes the type of animal, the breed of the animal, the `intake_type`, the `outcome_type`, and dates of these events. In addition, there is information about what condition the animal was in at the time of intake and at the time of the outcome.

The intentions with this data set is to answer a multitude of different questions that could help the animal shelter in the future. I think the following questions can help the animal shelter optimize a plan for incoming pets and try to maximize the opportunity for a pet to get adopted:

* Is there any spikes for adoptions and intake? And can I link them to any reasons?
* How long are animals typically at the shelter? How long for different outcomes?
* Which dog breeds are most likely to be adopted?
* Which dog breed is adopted the quickest?

# Data Description

To read in this data, I have downloaded the file from the above link and created a project to easily read in the file. This `R-Markdown` is part of a package. If you unzip and open the package, you should be able to run this code easily. One thing to note, the following packages were used to create this report and you will need to `install.packages` before running the code:

* `readxl` - to read in excel files.
* `scales` - for adjusting graphics legends and axes.
* `lubridate` - fast and user friendly way to work with dates.
* `rvest` - for scraping information off of the web.
* `RColorBrewer` - for changing graph colors
* `tidyverse` - a collection of packages to manipulate and graph data.


```{r library bank, echo=FALSE, message=FALSE}
#====library bank====
library(readxl)
library(scales, warn.conflicts = F)
library(lubridate,warn.conflicts = F)
library(rvest, warn.conflicts = F, quietly = T)
library(RColorBrewer)
library(tidyverse, warn.conflicts = F)
```

To get started, I will walk you through my analysis process. One thing I noticed immediately with this data set was that there were two tabs on the Excel sheet.
The first tab was labeled `raw`, while the second was labeled `simple`. To ensure I had all the data possible I did a quick comparison of the two data sets by `left_joining`  the `raw` tab's column names to the `simple` tab's column names and making sure there was no `NA` values. I then ensured the lengths of each tab were the same using `nrow`.

```{r col and row check, echo=FALSE}
#check that no columns in raw tab are in the simple tab.
suppressWarnings(data.frame(colname_simple = as.character(colnames(read_excel("data/week18_dallas_animals.xlsx", sheet = "simple")))) %>%
  left_join(data.frame(colname_simple = as.character(colnames(read_excel("data/week18_dallas_animals.xlsx", sheet = "raw"))),
                       colname_raw = as.character(colnames(read_excel("data/week18_dallas_animals.xlsx", sheet = "raw")))),
            by = "colname_simple"))

#check that rows are the same length
nrow(read_excel("data/week18_dallas_animals.xlsx", sheet = "simple")) == nrow(read_excel("data/week18_dallas_animals.xlsx", sheet = "raw"))
```

This proved that the `simple` tab is just a subset of the `raw` tab. Since the `raw` data is not overly large, I will just use the `raw` tab and filter out columns that really won't help. This way I will avoid missing any conclusions that can be drawn from the `raw` tab information that cannot be drawn from the `simple` tab. Below is a `glimpse` of the Dallas Animal Shelter columns which seemed potentially meaningful form the `raw` tab.

```{r read in file, echo=FALSE}
o_dallas_animals <- read_excel("data/week18_dallas_animals.xlsx", sheet = "raw", na = "NA")

o_dallas_animals <- o_dallas_animals %>%
  select(-c(kennel_status, tag_type, activity_number, activity_sequence, source_id,
            census_tract, receipt_number, impound_number, service_request_number,
            chip_status, year)) %>%
  rename(year = mo_year)

glimpse(o_dallas_animals)

```

# Data Manipulations

Below is a list of manipulations done in order for better analysis of the data:

* I added `intake_day` and `outcome_day` which represents the day of the week when the intake or outcome happened, respectively. This will allow for insight to activity on different days of the week.
* I added `total_time` which represents the number of days of the animal was at the shelter.
* I lumped the 20% least common types of dog breeds brought into the Dallas Animal Shelter as `OTHER`. During the analysis of this report, I discovered there were many different types of dog breeds brought into the shelter and many instance where there was only one or two dogs of that particular dog breed type which was brought into the shelter. Lumping the least common dog breeds into one breed type `OTHER` allows for better visual analysis of dog breeds. To do this, I used the following steps:
  + Counted dogs by breed type.
  + Arranged the breeds in descending order by count.
  + Added a column which gave the percent of total dogs.
  + Added a column which gave a cumulative percent of that percent column
  + Added a `new_breed` column which kept the original breed type of any animal under the 80% point of the cumulative percent column.

```{r manipulations, echo=FALSE}
dallas_animals <- o_dallas_animals %>%
  mutate(intake_day = wday(intake_date, label = TRUE),
         outcome_day = wday(outcome_date, label = TRUE),
         total_time = difftime(outcome_date, intake_date, units = "days"))

dog_breed <- dallas_animals %>%
  filter(animal_type == "DOG") %>%
  count(animal_breed) %>%
  rename(count_dogs = n) %>%
  arrange(desc(count_dogs)) %>%
  mutate(percent_dogs = round(count_dogs/sum(count_dogs), 2),
         cum_percent = cumsum(percent_dogs),
         new_breed = ifelse(cum_percent > 0.8, "OTHER", animal_breed))

dallas_animals <- dallas_animals %>%
  left_join(dog_breed %>%
              select(animal_breed, new_breed),
            by = "animal_breed")
```

# Analysis

The analsis is broken in down in three different "events". Intake, Time at Shelter, and Outcome. These sections represent the flow of an animal coming into the shelter, the time at the shelter, and then leaving the shelter.

## Intake

There is **`r format(nrow(dallas_animals), big.mark = ",", scientific = FALSE, trim = F)`** animals which the Dallas Animal Shelter has taken in. As you can see from the `Intake Time Series` graph below, there is about a years worth of data. The `Intake Time Series` graph also shows that the data from the Dallas Animal Shelter could have some seasonality. The smoothing line dips in the winter months and then spikes in the spring and summer months. However, one years worth of data is not really sufficient for identifying seasonality.

```{r time series intake, echo=FALSE, fig.cap="Intake Time Series"}
dallas_animals %>% count(intake_date) %>%
  ggplot(aes(x = intake_date, y = n)) +
  geom_line() +
  geom_smooth(method = 'loess') +
  ggtitle("Intake Time Series") +
  xlab("Intake Date") +
  ylab("Number of Animals") +
  theme(plot.title = element_text(face="bold")) +
  theme_bw()
```


The next thing I will look at is why animals are being administered to the Dallas Animal Shelter. The graph below shows different `intake_types` ordered from most common `intake_type` to least common. I can see that `Stray` and `Owner Surrender` are the highest on the list. This is expected. While there are going to be a few lost dogs and a few owners who surrender their dogs because of a conflict, most of them are likely to be found on the street.

```{r intake types, echo=FALSE, fig.cap="Intake Types"}
dallas_animals %>% count(intake_type) %>%   #count the number of animals
  ggplot(aes(x = reorder(intake_type, n), y = n)) +
  geom_bar(stat = "identity") +             #create a bar graph
  geom_text(aes(label = comma(n)),          #add labels to the graph with a 1,000's comma
            position = position_dodge(0.9),
            vjust = 0.1,
            hjust = -0.2) +
  ggtitle("Intake Types") +
  xlab("Intake Type") +
  ylab("Number of Animals") +
  scale_y_continuous(labels = comma, limits = c(0, 25000)) +
  theme(plot.title = element_text(face="bold")) +
  theme_bw() +
  coord_flip()
```

To look closer at these predominate intake types, I can take the column I added earlier, `intake_day`, which gives the day of the week of the `intake_date` and `filter` down to intake types equal to `STRAY` or `OWNER SURRENDER`. I can then make a graph to see if there are days of the week where an owner is more likely to surrender their pet or if it is more likely someone will call in or catch a stray.

```{r top intake by day, echo=FALSE, fig.cap="Intake Type by Day"}
dallas_animals %>% filter(intake_type == "STRAY" | intake_type == "OWNER SURRENDER") %>%
  ggplot(aes(x = intake_day, fill = intake_type)) +
  geom_bar() +
  facet_wrap( ~ intake_type) +
  ggtitle("Intake Type by Day") +
  xlab("Intake Day") +
  ylab("Number of Animals") +
  labs(fill = "Intake Type") +
  scale_y_continuous(labels = comma) +
  theme_bw() +
  theme(plot.title = element_text(face="bold")) +
  scale_fill_brewer(palette="Set1")
```

There are a couple things I can identify from the `Intake Type by Day` graph. One is that intake's are slow on Sundays. After reviewing the [Dallas Animal Shelter](https://dallascityhall.com/departments/dallas-animal-services/Pages/default.aspx)* hours online, the hours are shorter on Sunday. However, they are reasonable considering not many people work on Sunday (assuming I am looking at the right shelter). Another interesting, but unexplainable, conclusion that can be drawn from this graph, is that there is a larger number or strays found on Wednesdays.

*Caution: the animals are really cute on the above link!

Lastly, and probably most obvious while analyzing intake, I will look at what types of animals the Dallas Animal Shelter is taking in.

```{r animal types intake, echo=FALSE, fig.cap="Animal Intake Types"}
dallas_animals %>% count(animal_type) %>%   #count the number of animals
  ggplot(aes(x = animal_type, y = n)) +
  geom_bar(stat = "identity") +             #create a bar graph
  geom_text(aes(label = comma(n)),          #add labels to the graph with a 1,000's comma
            position = position_dodge(0.9),
            vjust = -0.3) +
  ggtitle("Animal Intake Types") +
  xlab("Animal Type") +
  ylab("Number of Animals") +
  scale_y_continuous(labels = comma) +
  theme(plot.title = element_text(face="bold")) +
  theme_bw()
```

As expected, there are way more dogs than anything else. Because of the large percentage of dogs relative to every other animal types, it will be worth taking a closer look at dogs specifically.
The below `Count of Dog Breeds` graph represents the number of each type of dog breed brought into the shelter.

```{r dog breed types, echo=FALSE, fig.cap="Count of Dog Breeds"}
dallas_animals %>%
  filter(animal_type == "DOG") %>%
  count(new_breed) %>%   #count the number of animals
  ggplot(aes(x = reorder(new_breed, n), y = n)) +
  geom_bar(stat = "identity") +             #create a bar graph
  geom_text(aes(label = comma(n)),          #add labels to the graph with a 1,000's comma
            position = position_dodge(0.9),
            vjust = 0.1,
            hjust = -0.2) +
  ggtitle("Count of Dog Breeds") +
  xlab("Dog Breed") +
  ylab("Number of Animals") +
  scale_y_continuous(labels = comma, limits = c(0, 6500)) +
  theme(plot.title = element_text(face="bold")) +
  theme_bw() +
  coord_flip()
```

From there I can create a graphic which tells me how the different types of dog breeds came into the shelter. To focus our analysis, I will `filter` down to the two most common intake types, `STRAY` and `OWNER SURRENDER`. Also, to put everything on the same scale I used the percent of dog breeds by `intake_type`. The graph below, `Dog Breed by Intake Types`, has the potential to show us what types of animals most commonly are surrendered by owner. Thankfully for humanity, but unfortunately for this graph, there is a large percent of each dog breed type that are brought in as strays. However, the `Dog Breed by Intake Types` graph does pull out some insight to what dog breeds that might be a little more difficult to deal with and are more likely to be brought in by the owner.

```{r dog breed intake, echo=FALSE, fig.cap="Dog Breed by Intake Types"}

dallas_animals %>%
  filter(animal_type == "DOG" & (intake_type == "STRAY" | intake_type == "OWNER SURRENDER")) %>%
  add_count(new_breed) %>%
  group_by(intake_type, new_breed) %>%
  summarise(percent_dogs = n()/mean(n)) %>%
  ggplot(aes(x = intake_type,
             y = new_breed,
             fill = percent_dogs)) +
  geom_tile() +
  ggtitle("Dog Breed by Intake Types") +
  xlab("Intake Type") +
  ylab("Animal Breed") +
  labs(fill = "Pct. of Dogs") +
  theme_bw() +
  theme(plot.title = element_text(face="bold"),
        axis.text.x = element_text(angle = 45)) +
  scale_fill_gradient(low = "white", high = "steelblue")
```

## Time at Shelter

For this section, I will look at the time animals spend at the shelter. An animal shelter obviously houses animals, but sadly, only has room for so many animals. To understand how long animals are in the shelter, I will use the variable I added in the beginning called `total_time` which indicates how many days the animal was in the shelter by using `datediff` on `intake_date` and `outcome_date`. I found that average amount of time an animal spends at the Dallas Animal Shelter is `r round(mean(dallas_animals$total_time, na.rm = T), 2)` days and the time average amount of time that a dog is there before being adopted is is 8.30 days. The table below shows average time at the animal shelter for different `outcome_types`. I can see that if there is an issue with a report, the animal will likely be there a longer time. It is interesting that there are outcome types with reporting issues and makes one wonder what actually physically happens to these animals.

```{r mean outcome type, echo=FALSE}
dallas_animals %>%
  group_by(outcome_type) %>%
  summarize(avg_time = round(mean(total_time, na.rm = T), 2)) %>%
  arrange(desc(avg_time)) %>%
  filter(outcome_type != "OTHER") %>%
  data.frame()
```

Next, I looked at the two most interesting and important outcomes, `ADOPTION` and `EUTHANIZED` in the `Histogram of Time at Shelter by Top Outcomes` graph. As you can see, the time spent at the Dallas Animal Shelter is usually not that long and extremely right skewed. However, the graphs align with our previous findings that animal spends about `r round(mean(dallas_animals$total_time, na.rm = T), 2)` at the shelter.

```{r time at shelter, echo=FALSE, fig.cap="Histogram of Time at Shelter by Top Outcomes"}
dallas_animals %>%
  filter(!is.na(total_time) & (outcome_type == "ADOPTION" | outcome_type == "EUTHANIZED")) %>%
  ggplot(aes(x = as.numeric(total_time), fill = outcome_type)) +
  geom_histogram(bins = 20, position = "dodge") +
  facet_wrap( ~ outcome_type) +
  ggtitle("Histogram of Time at Shelter by Top Outcomes") +
  xlab("Time at Shelter") +
  ylab("Number of Animals") +
  scale_y_continuous(labels = comma) +
  theme_bw() +
  theme(plot.title = element_text(face="bold"))
```

To bring the focus more to dogs again, the below `Average Time at Shelter Before Adoption` graph represents the average number of days that a particular dog breed spends at the Dallas Animal Shelter before it is adopted. While this animal shelter's data set does not have "animal size", but from basic knowledge of dog size, I can see that smaller dogs are adopted more quickly.

```{r dog breed out time, echo=FALSE, fig.cap="Average Time at Shelter Before Adoption"}
dallas_animals %>%
  filter(animal_type == "DOG" & !is.na(outcome_type) & outcome_type == "ADOPTION") %>%
  group_by(new_breed) %>%
  summarize(avg_time = as.numeric(mean(total_time, na.rm = T))) %>%
  ggplot(aes(x = reorder(new_breed, avg_time), y = avg_time)) +
  geom_bar(stat = "identity") +             #create a bar graph
  geom_text(aes(label = round(avg_time, 2)),          #add labels to the graph with a 1,000's comma
            position = position_dodge(0.9),
            vjust = 0.1,
            hjust = -0.2) +
  ggtitle("Average Time at Shelter Before Adoption") +
  xlab("Dog Breed") +
  ylab("Number of Days") +
  scale_y_continuous(limits = c(0, 12)) +
  theme(plot.title = element_text(face="bold")) +
  theme_bw() +
  coord_flip()
```

To further investigate this, I created an `adoption_ratio` for the `breed_type` which took the number of that breed that the shelter took in altogether and compared to those that were adopted of that breed. I then filter down to the top ten breed types and plotted the adoption ratio on the x-axis and number the number of days on the y axis to create the `Adoption Ratio by Adoption Days` graph below. I expected to see as the ratio went up, the amount of time it took to adopt the animal to go down. However, that doesn't appear to be the case. The amount of time before a dog gets adopted, again, apears to be more related to the size of dog than the likelyhood that it will be adopted.

```{r adoption ratio, echo=FALSE, fig.cap="Adoption Ratio by Adoption Days"}
dallas_animals %>%
  filter(animal_type == "DOG" & !is.na(outcome_date)) %>%
  add_count(new_breed, name = "count_dog_breed") %>%
  add_count(new_breed, outcome_type, name = "count_outcome_breed") %>%
  mutate(adopt_ratio = count_outcome_breed/count_dog_breed) %>%
  filter(outcome_type == "ADOPTION" & count_dog_breed >= 300) %>%
  ggplot(aes(x = adopt_ratio, y = as.numeric(total_time), color = new_breed))+
  geom_jitter() +
  ggtitle("Adoption Ratio by Adoption Days") +
  xlab("Adoption Ratio") +
  ylab("Number of Days") +
  labs(colour = "Dog Breed") +
  theme(plot.title = element_text(face="bold")) +
  theme_bw()

```

## Outcome

Next I will do the same type of time series analysis on the `outcome_date` as I did on the `income_date`. The `Outcome Time Series` graph below shows the same seasonality as our intake data, the smoothing line dips in the winter months and then spikes in the spring and summer months. While it is not as drastic of a sign for seasonality, this is not surprising since when it is warm out I want to be outside and walk with dogs, while some people could look at an animal being a burden in the winter time.

```{r time series outcome, echo=FALSE, fig.cap="Outcome Time Series"}
dallas_animals %>% filter(!is.na(outcome_date)) %>% count(outcome_date) %>%
  ggplot(aes(x = outcome_date, y = n)) +
  geom_line() +
  geom_smooth(method = 'loess') +
  ggtitle("Outcome Time Series") +
  xlab("Outcome Date") +
  ylab("Number of Animals") +
  theme(plot.title = element_text(face="bold")) +
  theme_bw()
```

Similar to our `Intake Time Series` graph, the `Outcome Time Series` graph has large spikes from day to day. I wanted to see if there was any effects of holidays on adoption. To do this, I used a federal holiday data set from [Kaggle](https://www.kaggle.com/gsnehaa21/federal-holidays-usa-19662020) and plotted them as vertical lines on our time series graph. From the below graph, `Adoption Around Holidays`, I can see that there are large spikes right after Thanksgiving and right before Christmas. This is expected after Thanksgiving because adoption centers usually have some sort of black Friday promotion and a puppy is obviously the perfect Christmas gift. I know I want one!

The other big spike before Labor Day is somewhat unexplained. My guess is that they had some sort of big promotion. I am pretty sure the humane society in my town does a promotion for free or $1 dogs at some point in the summer.

```{r holdays, echo=FALSE, fig.cap="Adoptoin Around Holidays"}
holidays <- read.csv("data/usholidays.csv", stringsAsFactors = F)
holidays <- holidays %>% select(Date, Holiday) %>% mutate(Date = ymd(Date)) %>%
  inner_join(dallas_animals %>%
               select(outcome_date) %>%
               mutate(outcome_date = ymd(outcome_date)),
              by = c("Date" = "outcome_date")) %>%
  distinct()

dallas_animals %>%
  filter(!is.na(outcome_date) & outcome_type == "ADOPTION") %>%
  count(outcome_date) %>%
  ggplot(aes(x = outcome_date, y = n)) +
  geom_line() +
  geom_smooth(method = 'loess') +
  ggtitle("Adoptoin Around Holidays") +
  xlab("Outcome Date") +
  ylab("Number of Animals") +
  theme_bw() +
  theme(plot.title = element_text(face="bold")) +
  geom_vline(aes(xintercept = as.POSIXct(Date)), colour = "magenta1", data = holidays) +
  ggrepel::geom_label_repel(aes(x = as.POSIXct(Date), label = Holiday),
                            colour = "grey37", y = 150, data = holidays)
```

Moving on, the `Outcome Types` graph below represents the `outcome_types` of animals from the Dallas Animal Shelter ordered from most common `outcome_types` to least common. I can see that `ADOPTION`, `EUTHENIZED`, and `TRANSFER` are the most common. `ADOPTION` and `EUTHENIZED` are expected to be the most common because that is the purpose of an animal shelter.

```{r outcome types, echo=FALSE, fig.cap="Outcome Types"}
dallas_animals %>% count(outcome_type) %>%   #count the number of animals
  ggplot(aes(x = reorder(outcome_type, n), y = n)) +
  geom_bar(stat = "identity") +             #create a bar graph
  geom_text(aes(label = comma(n)),          #add labels to the graph with a 1,000's comma
            position = position_dodge(0.9),
            vjust = 0.1,
            hjust = -0.2) +
  ggtitle("Outcome Types") +
  xlab("Outcome Type") +
  ylab("Number of Animals") +
  scale_y_continuous(labels = comma, limits = c(0, 12500)) +
  theme(plot.title = element_text(face="bold")) +
  theme_bw() +
  coord_flip()
```

Like I did in our `Dog Breed Intake Types` graph, I looked at percent of dog breeds by `outcome_type`. The `Dog Breed by Outcome Types` graph gives us insight to what types of animals most commonly are adopted or euthanized. As you can see, there are dog breeds which are more likely to be put down. These dogs are the breeds which someone would stereotype with being aggressive. For example, it is about a 50/50 chance that Pitbull's and Rottweiler's will be adopted or euthanized, so it is possible those stereo types are true. However, Pitbulls are the most common animal brought into the shelter so there is a bigger sample and it is possible they have a lesser standard for why they would euthanized.

```{r dog breed outcomes, echo=FALSE, fig.cap="Dog Breed by Outcome Types"}
dallas_animals %>%
  filter(animal_type == "DOG" & (outcome_type == "ADOPTION" | outcome_type == "EUTHANIZED")) %>%
  add_count(new_breed) %>%
  group_by(outcome_type, new_breed) %>%
  summarise(percent_dogs = n()/mean(n)) %>%
  ggplot(aes(x = outcome_type,
             y = new_breed,
             fill = percent_dogs)) +
  geom_tile() +
  ggtitle("Dog Breed by Outcome Types") +
  xlab("Outcome Type") +
  ylab("Dog Breed") +
  labs(fill='Pct. of Dogs') +
  theme_bw() +
  theme(plot.title = element_text(face="bold"),
        axis.text.x = element_text(angle = 45)) +
  scale_fill_gradient(low = "white", high = "steelblue")
```

To plan accordingly, it will be a good idea to look at the adoption rate on a daily basis. As you can see from the below `Adoption by Day` graph, Saturdays are the busiest days, which is not surprising. Another useful insight which can be seen from this graph is that the beginning of the work week is relatively busy compared to the end of the work week. With this information, the humane society can staff appropriately to speed up the process of customers adopting dogs and possibly increase the number of adoptionn as customers won't be waiting around deciding whether or not to adopt a dog.

```{r adoption by day, echo=FALSE, fig.cap="Adoption by Day"}
dallas_animals %>% filter(outcome_type == "ADOPTION" & !is.na(outcome_date)) %>%
  ggplot(aes(x = outcome_day)) +
  geom_bar() +
  ggtitle("Adoption by Day") +
  xlab("Day of Week") +
  ylab("Number of Animals") +
  scale_y_continuous(labels = comma) +
  theme_bw() +
  theme(plot.title = element_text(face="bold")) +
  scale_fill_brewer(palette="Set1")
```

# Conclusion

In conclusion, besides wanting to adopt a dog, I believe that the information provided, will be abe to help animal shelters better understand their data in multiple ways to save them time, money, and resources. I found that the Dallas Animal Shelter could have some seasonality in both their intake of dogs and outcome of dogs. I know that the most common intake type for the Dallas Animal Shelter is a stray and that strays are usually found on Wednesday's and most likely, a Pitbull. I found that there may be some dogs which are a little more difficult to deal with than other dogs due to the percent of owner surrenders, but there is no particular dog types which stick out as being extreamly difficult to deal with.

In our analysis when I examined how long animals are at the Dallas Animal Shelter I found that average amount of time an animal spends at the Dallas Animal Shelter is `r round(mean(dallas_animals$total_time, na.rm = T), 2)` days and the time average amount of time that a dog is there before being adopted is is 8.30 days and varys between dog breeds. Overally, I can see that animals are usually adopted withing a week or so, but even shorter amont of time if they are a small dog breed. In addtion to findings about adoption, I found that the Dallas Animal Shelter could possibly have a process or reporting problem which holds dogs up from being adopted.

When looking at the outcome of animals and dogs from the Dallas Animal shelter, I took a closer look at adoption through out the year and tied dates of adoption to holiday's. I found the dallas animal shelter can expect a lot more adoptions right after Thanksgiving and right before Christmas. I know that animals at the Dallas anmial shelter are likely going to be adopted, euthanized, or transfered and that more agressive dog breeds have a higher probability of being euthanized. In addtion, I know what the beginning of the week and Saturdays are usaully busier at the Dallas Animal Shelter.


# Plans for Further work:

To improve this study, I could look at things such as:

* Review promotions that the Dallas Animal Shelter has by scraping their facebook or twitter.
* Include how dog size relates to adoption by finding a data set with dog size and then cleaninng and standarizing the dog names and joining to our data set.
* When I `glimpsed` at the data you could see there were kennel numbers. If I had enough data I could see if location of animals and adoption rates are related.
* I could break down adoption by month to see if there is busier months than others, or even times to see if the selter is busier at certain times to staff apporpiately.
